---
title: "Prediction of excercise quality based on personal activity data"
author: "Julia Hoffman"
date: "5/9/2017"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(caret)
library(cvTools)
library(gbm)
library(plyr)
library(randomForest)
```

## Summary

With the use of devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity. These devices are part of the quantified self movement – people who take measurements about themselves regularly to improve their health and/or to find patterns in their behavior. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants to predict how well they do the excercises. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

More information about this data is available from the website here: <http://groupware.les.inf.puc-rio.br/har> (section on the Weight Lifting Exercise Dataset). The data for this project also come from this site.

## Data Preprocessing

```{r, load_data}
trainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "pml-training.csv"
testFile  <- "pml-testing.csv"

if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile=trainFile, method="curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile=testFile, method="curl")
}

training <- read.csv(file = trainFile, header = TRUE, sep = ",")
testing  <- read.csv(file = testFile, header = TRUE, sep = ",")

dim(training); dim(testing)
```

The training file contains 19622 observations of 160 variables, while the test 
file has 20 observations of 160 variables.
Closer look at training data reveals that there are a lot of NA values in a 
number of columns. We check now how many rows in each column are affected, to 
see if it's better to remove observations, or the variables.

```{r, na_check}
training_na <- training[, colSums(is.na(training)) != 0]
na_count <-sapply(training_na, function(na) sum(is.na(na)))
na_count <- data.frame(na_count)
```

The 67 columns that have NA values contain almost only NA values (this check is 
not shown here). In each case 19216 observations (98% of the training sample) are NA, therefore these variables will be deleted. Additionally, first 7 colums of data are not useful for predictions, they will be removed as well. Out of remaining variables there are still a few which are empty, they will also be removed. This leaves the training sample with 53 predictors.

```{r, cleaning}
trainingData <- training[, colSums(is.na(training)) == 0]#removing variables with NA
trainingData <- trainingData[ ,-c(1:7)]#removing irrelevant/empty variables
classe <- trainingData$classe
trainingData <- trainingData[, sapply(trainingData, is.numeric)]
trainingData$classe <- classe

testingData <- testing[, colSums(is.na(testing)) == 0]
testingData <- testingData[ ,-c(1:7)]
testingData <- testingData[, sapply(testingData, is.numeric)]

dim(trainingData); dim(testingData)
```

The training set will be divided now into train and validation samples, to help
model selection.

```{r, train_n_test}
inTrain <- createDataPartition(y = trainingData$classe, p = 0.7, list = FALSE)
trainSub <- trainingData[inTrain,]
validSub <- trainingData[-inTrain,]
```

## Prediction model evaluation

We consider here classification trees (`rpart`), random forests (`rf`), and 
boosted regression (`gbm`) models. The cross validation is done by `cv` method 
in the trainControl, for which 5-fold method was chosen.

#### Classification trees (rpart)

```{r, rpart}
set.seed(123)
trCtrl <- trainControl(method = 'cv', number = 5, summaryFunction = defaultSummary)
grid_rpart <- expand.grid(cp = seq(0, 0.05, 0.005))#complexity parameter
fit_rpart <- train(classe~., data = trainSub, method = 'rpart', trControl = trCtrl,tuneGrid = grid_rpart)
print(fit_rpart)
plot(fit_rpart)
###
folds = createFolds(trainSub$classe, k = 5)
error = rep(0, 5)
for (k in 1:5) {
    test = trainSub[unlist(folds[k]), ]
    pred = predict(fit_rpart, test)
    accuracy = sum(as.numeric(pred) == as.numeric(test$classe)) / length(pred)
    error_k = 1 - accuracy
    error[k] = error_k
}
error_rpart = round(mean(error)*100, digits = 2)
```

The out of sample error estimations for classification tree is **`r paste0(error_rpart,"%")`**. This is the weakest model of all considered.

#### Boosted regression (gbm)

```{r, gbm}
set.seed(456)
trCtrl <- trainControl(method = 'cv', number = 5, summaryFunction = defaultSummary)
grid_gbm <- expand.grid( n.trees = seq(50, 200, 5), interaction.depth = c(10), shrinkage = c(0.1), n.minobsinnode = 20)
fit_gbm <- train(classe~., data = trainSub, method = 'gbm', trControl = trCtrl, tuneGrid = grid_gbm, verbose = FALSE)
print(fit_gbm)
plot(fit_gbm)
###
folds = createFolds(trainSub$classe, k = 5)
error = rep(0, 5)
for (k in 1:5) {
    test = trainSub[unlist(folds[k]), ]
    pred = predict(fit_gbm, test)
    accuracy = sum(as.numeric(pred) == as.numeric(test$classe)) / length(pred)
    error_k = 1 - accuracy
    error[k] = error_k
}
error_gbm = round(mean(error)*100, digits = 7)
```

The out of sample error estimations for boosted regression is **`r paste0(error_gbm,"%")`**. This model has much higher accuracy with a very small (near zero) error .

#### Random forests (rf)

```{r, rf}
set.seed(789)
trCtl <- trainControl(method = 'cv', number = 5, summaryFunction = defaultSummary)
grid_rf <- expand.grid( mtry = seq(2, 40, 5))
fit_rf <- train(classe~., data = trainSub, method = 'rf', trControl = trCtl, tuneGrid = grid_rf, verbose = FALSE)
print(fit_rf)
plot(fit_rf)
###
folds = createFolds(trainSub$classe, k = 5)
error = rep(0, 5)
for (k in 1:5) {
    test = trainSub[unlist(folds[k]), ]
    pred = predict(fit_rf, test)
    accuracy = sum(as.numeric(pred) == as.numeric(test$classe)) / length(pred)
    error_k = 1 - accuracy
    error[k] = error_k
}
error_rf = round(mean(error)*100, digits = 7)
```

The out of sample error estimations for random forests is **`r paste0(error_rf,"%")`**. This model also has high accuracy and low (near zero) error.
Let's compare boosted regression and random forests models, as they both perform 
better than classification tree (rpart).

```{r, plots}
results <- resamples(list(GBM = fit_gbm, RandomForest = fit_rf))

# summarize the distributions
summary(results)
bwplot(results)
```

## Model choice and prediction

Model comparison suggests that among checked models, boosted regression perform 
as well as random forests, with very slight differences. The random forests model 
was chosen as the final model. 

One can check the model accuracy on full (not folded) trainSub sample and 
check out of sample error on the validSub sample:

```{r,final_training}
set.seed(333)
grid <- expand.grid(fit_rf$bestTune)
model <- train(classe~., data = trainSub, method = 'rf', tuneGrid = grid, verbose = FALSE)

prediction <- predict(model, validSub)
error = 1- sum(as.numeric(prediction) == as.numeric(validSub$classe)) / length(prediction)
```

The training model accuracy is **`r paste0(round(confusionMatrix(validSub$classe, prediction)$overall[1]*100,2),"%")`** and the out of sample error on validSub data set for this model is **`r paste0(round(error*100,2),"%")`**

## Test sample prediction

The final test sample prediction gives the following results:

```{r,final_prediction}
set.seed(3234)
final_prediction <- predict(model, testingData)
final_prediction
```